Extended Data

### 2021-08-23 doc/manuscript/Dow_Extended_Data.Rmd
remove Fig. 3 from main manuscript
(as planned in our meeting)

Co-Authored-By: Cameron Dow <55721987+camerondow35@users.noreply.github.com>

#64

pheno_Tmax_climwin_caption <- "**Extended Data Figure 4 | Relationship between growth parameters and mean maximum temperature within the critical temperature window (CTW).** CTW was defined using *climwin*. For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59). CAPTION NEEDS MORE DETAIL**)" #needs attention!!!

>>>>

pheno_Tmax_climwin_caption <- "**Extended Data Figure 4 | Response of the timing of stem growth phenology to spring temperatures.** The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declined significantly with mean $T_{max}$ during their respective critical temperature window (CTW). For each CTW $T_{max}$, the posterior mean of the fitted day of year is represented by the solid line and 95% credible intervals are represented by bands. (**[FIGURE NEEDS SOME WORK/ MORE INFO](https://github.com/EcoClimLab/growth_phenology/issues/59). CAPTION NEEDS MORE DETAIL**)" #needs attention!!!

### 2021-08-23 doc/manuscript/Dow_manuscript_Nature.Rmd
remove Fig. 3 from main manuscript
(as planned in our meeting)

Co-Authored-By: Cameron Dow <55721987+camerondow35@users.noreply.github.com>

#64

#Figure numbers
schematic_summary=1 #illustrates parameters
DOY_timing = 2
  DOY_timing_SCBI = "2a"  #currently 4 panel!!!!
  DOY_timing_Harvard = "2b"  #currently 4 panel!!!!
pheno_Tsensitivity = 3 
tree_rings = 4

@@

DOY_Tsensitivity_caption <- "**Figure 3 | Response of the timing of stem growth phenology to spring temperatures.** The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declined significantly with mean $T_{max}$ during their respective critical temperature window (CTW). For each CTW $T_{max}$, the posterior mean of the fitted day of year is represented by the solid line and 95% credible intervals are represented by bands. Relationships for additional growth parameters (Fig. 1a) are given in Extended Data Figure 4. (**DOYrmax will be dropped.** [GitHub Issue 23](https://github.com/EcoClimLab/growth_phenology/issues/23))" #needs attention!!!

tree_rings_caption <- "**Figure 4 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for 207 chronologies from 114 sites across eastern North America** (Extended Data Figure 1). Chronologies are grouped by xylem porosity and ordered by mean maximum April temperature. Chronology details are given in the Supplementary Information. (NOTE: Figure still needs some work. **COLOR SCALE IS  A BIT OFF IN THIS COMPOSITE FIGURE (scales not identical across panels).**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)" #needs attention!!! #info subject to change!!!

@@

Both leaf phenology and stem growth were sensitive to spring temperatures (Figs. `r DOY_timing` -`r pheno_Tsensitivity`). 
While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-7), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan. 1 and $DOY_{25}$  for which the correlation between $T_{max}$ and $DOY_{25}$ was strongest (henceforth, critical temperature window, CTW; Fig. `r schematic_summary`). 
The strongest relationships, including but not limited to the CTW (`r  EDT_mean_growth_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$ (`r EDF_climwin`).

![`r DOY_Tsensitivity_caption`](tables_figures/DOY_Tsensitivity.png){ width=50% }

For both ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`-`r pheno_Tsensitivity`).
Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced significantly with increasing $T_{max}$ during the CTW (p=## and ##, respectively). 
Specifically, [greenup DOY occurred 21 days earlier at SCBI and 7 days earlier at Harvard Forest when comparing the years with the warmest and coldest CTW $T_{max}$](https://github.com/EcoClimLab/growth_phenology/issues/58), and mid-greenup also advanced, but to a lesser extent (Fig. `r DOY_timing`).
Similarly, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at both sites all decreased significantly with CTW $T_{max}$, advancing [1-8 days per degree Celsius](https://github.com/EcoClimLab/growth_phenology/issues/58) during the CTW (Figs. `r schematic_summary`- `r pheno_Tsensitivity`, `r DOY_timing`; see also Extended Data Figures **4-6**).

Whereas the length of time between leaf greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer springs (Fig. `r DOY_timing`), there was no consistent lengthening of the period of peak stem growth. 
Specifically, because $DOY_{75}$ -- and not only $DOY_{25}$ -- shifted earlier in years with warm springs, $L_{PGS}$ displayed mixed responses to spring temperatures (Fig. `r schematic_summary`, Extended Data Figures **4-6**).

In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were less consistent and generally quite subtle (Figs. `r schematic_summary`, `r pheno_Tsensitivity`). 
Neither maximum seasonal growth rate, $g_{max}$, nor annual growth measured with dendrometer bands, $\Delta DBH$, was consistently linked to early spring temperatures (Fig. `r schematic_summary`, Extended Data Figures **4-6**).
Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_mean_growth_parameters`), displayed either no relationship to CTW $T_{max}$ (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest),
while $\Delta DBH$ displayed no relationship with CTW $T_{max}$ (`r EDF_pheno_Tmax_climwin`). 
Given the subtlety of these responses, a much longer record was needed to clarify what, if any, effect spring temperatures have on stem growth rates.

@@

Despite remaining uncertainties, our findings contribute substantially to a broader picture of the influence of spring temperatures on forest ecosystem function. 
Warm spring temperatures result in earlier leaf-out (Fig. `r DOY_timing`)[@jeong_phenology_2011], an earlier start to net C sequestration[@keenan_net_2014], and, as shown here, earlier initiation of stem growth (Figs. `r schematic_summary`- `r pheno_Tsensitivity`).
Yet, an earlier start to the growing season is not associated with a longer growing season or increased growth (Figs. `r schematic_summary`, `r tree_rings`), presumably because stem diameter growth is curtailed by C sink limitation[@zani_increased_2020] or drought[@dorangeville_drought_2018;@helcoski_growing_2019].

>>>>

#Figure numbers
schematic_summary=1 #illustrates parameters
DOY_timing = 2
  DOY_timing_SCBI = "2a"  #currently 4 panel!!!!
  DOY_timing_Harvard = "2b"  #currently 4 panel!!!!
tree_rings = 3

@@

tree_rings_caption <- "**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for 207 chronologies from 114 sites across eastern North America** (Extended Data Figure 1). Chronologies are grouped by xylem porosity and ordered by mean maximum April temperature. Chronology details are given in the Supplementary Information. (NOTE: Figure still needs some work. **COLOR SCALE IS  A BIT OFF IN THIS COMPOSITE FIGURE (scales not identical across panels).**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)" #needs attention!!! #info subject to change!!!

@@

Both leaf phenology and stem growth were sensitive to spring temperatures (Figs. `r DOY_timing`, `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`). 
While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-7), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan. 1 and $DOY_{25}$  for which the correlation between $T_{max}$ and $DOY_{25}$ was strongest (henceforth, critical temperature window, CTW; Fig. `r schematic_summary`). 
The strongest relationships, including but not limited to the CTW (`r  EDT_mean_growth_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$ (`r EDF_climwin`).

For both ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], at our focal sites the DOYs of leaf greenup and mid-greenup (measured at ecosystem level via remote sensing) advanced significantly with increasing $T_{max}$ during the CTW (p=## and ##, respectively). 
Specifically, [greenup DOY occurred 21 days earlier at SCBI and 7 days earlier at Harvard Forest when comparing the years with the warmest and coldest CTW $T_{max}$](https://github.com/EcoClimLab/growth_phenology/issues/58), and mid-greenup also advanced, but to a lesser extent (Fig. `r DOY_timing`).
Similarly, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ at both sites all decreased significantly with CTW $T_{max}$, advancing [1-8 days per degree Celsius](https://github.com/EcoClimLab/growth_phenology/issues/58) during the CTW (Figs. `r schematic_summary`, `r DOY_timing`; see also Extended Data Figures `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`).

Whereas the length of time between leaf greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer springs (Fig. `r DOY_timing`), there was no consistent lengthening of the period of peak stem growth. 
Specifically, because $DOY_{75}$ -- and not only $DOY_{25}$ -- shifted earlier in years with warm springs, $L_{PGS}$ displayed mixed responses to spring temperatures (Fig. `r schematic_summary`, Extended Data Figures `r EDF_pheno_Tmax_climwin`-`r EDF_pheno_Tmin_AprilMay`).

In contrast to the pronounced effects of $T_{max}$ on the timing of growth, its effects on $g_{max}$ and $\Delta DBH$ were less consistent and generally quite subtle (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`). 
Neither maximum seasonal growth rate, $g_{max}$, nor annual growth measured with dendrometer bands, $\Delta DBH$, was consistently linked to early spring temperatures (Fig. `r schematic_summary`, Extended Data Figures **4-6**).
Specifically, $g_{max}$, which occurred very close to $DOY_{50}$ (on $DOY_{g.max}$; `r EDT_mean_growth_parameters`), displayed either no relationship to CTW $T_{max}$ (SCBI), or extremely small changes in opposite directions for ring- and diffuse- porous species (Harvard Forest),
while $\Delta DBH$ displayed no relationship with CTW $T_{max}$ (`r EDF_pheno_Tmax_climwin`). 
Given the subtlety of these responses, a much longer record was needed to clarify what, if any, effect spring temperatures have on stem growth rates.

@@

Despite remaining uncertainties, our findings contribute substantially to a broader picture of the influence of spring temperatures on forest ecosystem function. 
Warm spring temperatures result in earlier leaf-out (Fig. `r DOY_timing`)[@jeong_phenology_2011], an earlier start to net C sequestration[@keenan_net_2014], and, as shown here, earlier initiation of stem growth (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
Yet, an earlier start to the growing season is not associated with a longer growing season or increased growth (Figs. `r schematic_summary`, `r tree_rings`), presumably because stem diameter growth is curtailed by C sink limitation[@zani_increased_2020] or drought[@dorangeville_drought_2018;@helcoski_growing_2019].
